Efficiently summarising relationships in large samples: a general duality between statistics of genealogies and genomes

### 2019-09-03 treestats_paper.tex
Algorithm description.

\newcommand{\Nt}{\mathcal{N}}  % node table
\newcommand{\Et}{\mathcal{E}}  % edge table
\newcommand{\prop}[1]{.\mbox{\texttt{#1}}} % property of a thing
\newcommand{\indexin}[0]{\ensuremath{\mathcal{I}}}
\newcommand{\indexout}[0]{\ensuremath{\mathcal{O}}}

\newcommand{\plr}[1]{{\color{blue}\textbf{plr:} \it #1}}
\newcommand{\jk}[1]{{\color{red}\textbf{jk:} \it #1}}

@@

%%%%%%%%%%%%%%%%%%%%%
\subsection*{Algorithms}
In this section we provide describe and analyse the algorithm used
to compute branch length statistics. This is a generalation
of the algorithm used to maintain the number of samples from a
given set that derive from each node in a tree sequence.
\cite[Algorithm L]{kelleher2016efficient}.

\jk{Plan for section}
\begin{itemize}
\item First paragraph is a general outline of the strategy.
\item Explanation of notation and the index vectors (we should probably define
these again, in the interest of completeness. I'm not sure the definition
in the msprime paper really worked)
\item Algorithm listing.
\item Next para is an explanation of the algorithm in words, referring
to the figure for what we mean by adding and removing edges.
\item Next para is an analysis of the algorithm. We note here that
we propagate stuff all the way up the tree close to optimally, because of the
ordering of edge insertions and removals. For example, we always disconnect
a subtree from the main tree first, meaning that the only time we go all the
way to the root is the first disconnection.
Discuss the figure perf figure.
\end{itemize}

\plr{Question: Figures 1, 2 and 7 show the same sort of thing
    (fig 1 is a site stat, aiming to give the general overview;
    figs 2 and 7 are branch stats). Should they be more similar?
    Do we need all of these?}

\plr{right now things like $\Et\mbox{.right}$ read like a funnily terminated sentence.
     I've changed these with a macro, feel free to adjust.}

\clearpage
\begin{taocpalg}{B}{General branch statistics}
{Given a list of $n$ sample nodes $S$, corresponding weights $w$ and
summary function $f: \R \rightarrow \R$
compute the span-normalised statistic $\sigma$
over a tree sequence with sequence length $L$
defined by the node and edge tables $\Nt$ and $\Et$. We assume
that the index vectors $\indexin$ and $\indexout$ have been precomputed.
\plr{I assume you plan to say in words what $\beta$, $x$, $F$, etc are}
}

@@

\algstep{B2.}{Terminate.}{If $j \geq |\Et|$ or $t_l \geq L$
    set $\sigma \leftarrow \sigma / L$ and terminate.
    \plr{perhaps say "return $\sigma/L$" to make it more clear what the output is?}
}

@@

\algstep{B5.}{Propagate loss.}{While $v \neq -1$, set
    $s \leftarrow s - \beta_v F_v$, $x_v \leftarrow x_v - x_u$,
    $F_v \leftarrow f(x_v)$, $s \leftarrow s - \beta_v F_v$
    and $v \leftarrow \pi_v$. Afterwards, go to \algref{B3}.
}

@@

\algstep{B8.}{Propagate gain.}{While $v \neq -1$, set
    $s \leftarrow s - \beta_v F_v$, $x_v \leftarrow x_v + x_u$,
    $F_v \leftarrow f(x_v)$, $s \leftarrow s - \beta_v F_v$
    and $v \leftarrow \pi_v$. Afterwards, go to \algref{B6}.
}

\algstep{B9.}{Tree loop tail.}{Set $t_r \leftarrow L$.
    If $j < |\Et|$ set $t_r \leftarrow \min(t_r, \Et[\indexin_j]\prop{left})$.
    Then, if $k < |\Et|$ set $t_r \leftarrow \min(t_r, \Et[\indexout_k]\prop{right})$
    Finally, set $\sigma \leftarrow (t_r - t_l)s$,  $t_l \leftarrow t_r$
    and go to \algref{B2}.
    \plr{I think you mean to add to $\sigma$, not replace it?}
}

@@

@@

\caption{
Illustration of the internal state for algorithm B for $\iw = \bone_S$ and
$f(x) = x(5 - x) / 20$ as in Example~\ref{ex:site_diversity}. (A)
Immediately before we remove the edge joining 6 to 7 and (B) immediately after.
See the text for descriptions of the arrays encoding the state.
\label{fig:algorithm_example}}
\end{figure}

>>>>

\newcommand{\Nt}{\mathcal{N}}  % node table
\newcommand{\Et}{\mathcal{E}}  % edge table
\newcommand{\prop}[1]{.\mbox{\texttt{#1}}} % property of a thing
% Not sure about this notation. Want to use i and o, but using normal
% math fonts would make things less clear, I think.
\newcommand{\indexin}[0]{\ensuremath{\mathbf{i}}}
\newcommand{\indexout}[0]{\ensuremath{\mathbf{o}}}
\newcommand{\plr}[1]{{\color{blue}\textbf{plr:} \it #1}}
\newcommand{\jk}[1]{{\color{red}\textbf{jk:} \it #1}}

@@

%%%%%%%%%%%%%%%%%%%%%
\subsection*{Algorithms}
In this section we describe and analyse the algorithm used
to compute branch length statistics. This is a generalization
of the algorithm used to maintain the number of samples from a
given set that derive from each node in a tree
sequence \cite[Algorithm L]{kelleher2016efficient}. The central
problem which is shared across site, branch and node statistics
is to efficiently maintain the ``state'' $x_u$ (where $x_u$ is the
sum of the weights for all samples descending
from, and including, node $u$; see Figure~\ref{fig:divergence_diagram}).
As we move along the sequence, the trees change when branches are inserted
and removed. By carefully ordering these insertions and removals, we
ensure that this state is always correctly maintained and that we
only need to make small adjustments for each tree. Computing the statistics
is then a straightforward matter of applying the summary function $f$
to the node states and aggregating the results appropiately for the
particular statistics mode and windowing options.

Formally, we define a tree sequence topology via a set of
tables~\citep{kelleher2018efficient}. In a tree sequence, a node
describes a specific ancestral haplotype, and details about these
ancestors are stored in a node table, $\Nt$. The row $\Nt[u]$
contains all information about the node $u$, and rows
are indexed from zero such that $0 \leq u < |\Nt|$.
In particular, the birth time for $u$ is given by
$\Nt[u]\prop{time}$. Information about how nodes relate to each
other along the genome is encoded in an edge table, $\Et$.
For a given row $j$, $\Et[j]\prop{child}$ and $\Et[j]\prop{parent}$
define a parent-child relationship between two nodes in a set of
contiguous trees along the genome. $\Et[j]\prop{left}$ then denotes
the left-most (inclusive) and $\Et[j]\prop{right}$ the right-most
(exclusive) genome coordinates over which this branch exists.
The order in which edges are inserted and removed is determined
by the ``index vectors'' $\indexin$ and $\indexout$~\citep{kelleher2016efficient},
which are defined as follows. \jk{TODO: succinct definition of these}.
As we move along the tree sequence, the topology is recorded in the
parent vector $\pi$, such that $\pi_u$ is the parent of node $u$
and $\pi_u = -1$ if $u$ is a root. Similarly, we maintain the
branch length $\beta$ such that
$\beta_u = \Nt[\pi_u]\prop{time} - \Nt[u]\prop{time}$ and
$\beta_u = 0$ if $u$ is a root. As evaluating the summary function
$f$ is an expensive operation, we also maintain the vector $F$
such that $F_u = f(x_u)$. For notational simplicity here we assume
that weights and summary functions are scalars, but the extension to
vectors is trivial. We also assume that we are computing the statistic
for across the full span of the tree sequence; the extension to multiple
windows is routine.

\begin{taocpalg}{B}{General branch statistics}
{Given a list of $n$ sample nodes $S$, corresponding weights $w$ and
summary function $f: \R \rightarrow \R$
compute the span-normalised statistic $\sigma$
over a tree sequence with sequence length $L$
defined by the node and edge tables $\Nt$ and $\Et$. We assume
that the index vectors $\indexin$ and $\indexout$ have been precomputed.
}

@@

\algstep{B2.}{Terminate.}{If $j \geq |\Et|$ or $t_l \geq L$,
    return $\sigma / L$ and terminate.
}

@@

\algstep{B5.}{Propagate loss.}{While $v \neq -1$, set
    $s \leftarrow s - \beta_v F_v$, $x_v \leftarrow x_v - x_u$,
    $F_v \leftarrow f(x_v)$, $s \leftarrow s + \beta_v F_v$
    and $v \leftarrow \pi_v$. Afterwards, go to \algref{B3}.
}

@@

\algstep{B8.}{Propagate gain.}{While $v \neq -1$, set
    $s \leftarrow s - \beta_v F_v$, $x_v \leftarrow x_v + x_u$,
    $F_v \leftarrow f(x_v)$, $s \leftarrow s + \beta_v F_v$
    and $v \leftarrow \pi_v$. Afterwards, go to \algref{B6}.
}

\algstep{B9.}{Tree loop tail.}{Set $t_r \leftarrow L$.
    If $j < |\Et|$ set $t_r \leftarrow \min(t_r, \Et[\indexin_j]\prop{left})$.
    Then, if $k < |\Et|$ set $t_r \leftarrow \min(t_r, \Et[\indexout_k]\prop{right})$
    Finally, set $\sigma \leftarrow \sigma + (t_r - t_l)s$,  $t_l \leftarrow t_r$
    and go to \algref{B2}.
}

@@

Algorithm \algref{B} begins by setting the initial state for $\pi$, $\beta$,
$x$ and $F$ for each node in the tree sequence, and then sets the values
of $x_u$ and $F_u$ for each of the samples in $S$. Afterwards, we set
our ``running sum'' $s$ and output statistic $\sigma$ to zero, along
with the ``tree left'' variable $t_l$. The $j$ and $k$ variables are
used to keep track of our position in the edge insertion and edge
removal indexes, respectively. After completing initialisation in
\algref{B1}, we enter the main tree loop in \algref{B2} which is
run once for each tree in the sequence. As we are processing each
tree, we keep track of the edges that need to be processed with
the $t_l$ variable, which stores the left-most genome coordinate
in the current tree. The first thing we do for a new tree is to
remove any edges that were in the previous tree are not in the
current tree. These must be the edges in which the right coordinate
is equal to $t_l$, and so \algref{B3} loops over these edges using
the edge removal index $\indexout$. Step \algref{B4} then removes
the branch $u \mapsto v$ corresponding to a single edge, by
subtracting its contribution to the running sum $s$, setting the parent of
$u$ to $-1$ and its branch length to $0$. As shown in
Figure~\ref{fig:algorithm_example}B, removing an edge will result
in the subtree rooted at $u$ become disconnected from the rest of the
tree. Step \algref{B5} ensures that the state of the rest of the tree
is correctly maintained by propagating the loss of the state $x_u$
up the tree from $v$. Thus, for each direct ancestor $v$ that was a
direct ancestor of $u$ in the previous tree, we first subtract
its contribution from the running sum $s$ and remove the contribution
of $x_u$ from its state. Since the state of $v$ has changed, we must
recompute the value of our summary cache $F_v$, and then finally
add the new contribution of $v$ to $s$. (For example, note the changes in $x$
and $F$ for nodes 7 and 8 in Figure~\ref{fig:algorithm_example}B.)

Once all of the edges with right coordinate equal to $t_l$ have been removed in
steps \algref{B3--B5}, we then insert edges that start in the current tree
(i.e., with left coordinate $t_l$) in steps \algref{B6--B8}. We update the
state to account for the new branch $u \mapsto v$ by setting the parent of
$u$ to $v$, computing the branch length of $u$ and adding the new contribution
of $u$ to $s$ in \algref{B7}. Then, as adding a new edge can connect the
subtree rooted at $u$ to the larger tree at $v$, we propagate the gain of $x_u$
up the tree in step \algref{B8} (note the symmetry with step \algref{B5}).
Finally, once we have remove and inserted all of the relevant edges, we are
ready to add the contribution of the current tree to the overall statistic,
$\sigma$ and move on to the next tree in step \algref{B9}. To do this, we first
compute the right-hand coordinate of the tree $t_r$ (a process slightly
complicated by the possible presence of ``gaps'' in the tree sequence, spanned
by no edges). Then, we add the running sum $s$ weighted by the span of the
current tree $t_r - t_l$ to $\sigma$, and return to \algref{B2} to process the
next tree. If we have reached the end of the sequence, we then divide $\sigma$
by the sequence length to renormalise and exit.

To analyse Algorithm~\algref{B} we will assume that the tree sequence has
been simulated under the standard coalescent with a sample of size
$n$ and population scaled recombination rate $\rho$. Under these conditions,
we know that there are $O(n + \rho \ln n)$ edges in the tree
sequence~\cite{kelleher2016efficient}. Clearly, each edge is examined
exactly once in steps \algref{B6} and \algref{B7} in order to create
the trees, and at most once in steps \algref{B3} and \algref{B4} (we do
not remove the edges for the final tree). Additionally, each edge that
we insert or remove may incur the cost of traversing up the tree as
far as the root in steps \algref{B5} and \algref{B8}.
As coalescent genealogies are asymptotically
balanced~\citep{li2013coalescent}, the expected number of nodes
on the path to root is $\log_2 n$. Therefore, the overall running time
of Algorithms~\algref{B} is
\[
O(n \log_2 n + \rho \log_2 n \ln n)
\]
\jk{Slightly iffy on what this should be in $O$ terms. I'm pretty
sure that the total number of edges of log base $e$ and the
expected height is log base 2. Not sure what the asymptotics
say here.}
The asymptotic analysis here assumes that we traverse upwards to root
for each edge, but this is not the case. Edges are removed in
oldest-first order, guaranteeing that the deepest node any subtree being
modified is processed first. Therefore, this subtree will be disconnected from the
main tree, ensuring that we traverse upwards all the way to root once. Similarly,
edges are inserted in youngest-first order, ensuring that subtrees are only
reconnected to the main tree when they are fully built.

@@

\caption{
Illustration of the internal state for algorithm B for $\iw = \bone_S$ and
$f(x) = x(5 - x) / 20$ as in Example~\ref{ex:site_diversity}. (A)
Immediately before we remove the edge joining 6 to 7 and (B) immediately after.
See the text for descriptions of the arrays encoding the state.
\jk{TODO: the value of $s$ would be helpful here too.}
\label{fig:algorithm_example}}
\end{figure}
